Report on a Safety Checker for Petri nets

### 2013-11-25 paper/workspace/main.tex
Updated table of experiments in paper draft.

\title{Report on a Safety Checker for Petri nets}
\author{Philipp Meyer \and Rusl\'{a}n Ledesma-Garza}
% \institute{Technische Universit\"at M\"unchen}
\date{Wed Nov 13 09:11:15 CEST 2013}

@@

\begin{center}
  \begin{tikzpicture}
    \node[state] (begin) {BEGIN};
    \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
    \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
    \node[print, right=of satc] (yes) {YES};
    \node[print, below=of satc] (dontknow) {Don't know};
    \node[state, right=of yes] (end1) {END};
    \node[state, below=of dontknow] (end2) {END};

@@

  Subprocedure TrapConditions constructs trap conditions C_\theta corresponding to N and A.
  Subprocedure \Delta constructs refinement constraint \delta corresponding to A_\theta.
\end{verbatim}
\begin{center}
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
  \node[action, below=of satc] (modelc) {$A:=\text{Model}(C \cup \{\neg P\})$\\
                               	$C_{\theta}:=\text{TrapConditions}(N, A)$};
  \node[decision, below=of modelc] (satctheta) {$\text{SAT}(C_\theta)$};
  \node[action, below=of satctheta] (modelctheta) {$A_\theta:=\text{Model}(C_\theta)$\\
                                    	$\delta:=\Delta(A_\theta)$\\
                                    	$C:=C \cup \{\delta\}$};
  \node[print, right=of satc] (yes) {YES};
  \node[print, right=of satctheta] (dontknow) {Don't know};
  \node[state, right=of yes] (end1) {END};
  \node[state, right=of dontknow] (end2) {END};

@@

We experimented on the benchmarks \ttt{cprover-PN} and
\ttt{cprover\_software\_analysis}
available at
\ttt{http://www.mpi-sws.org/\~{}jkloos/iic-experiments/cprover.zip}.
The set of benchmarks consists 77 benchmarks, 26 positive and 51 negative.
We applied 8 tool configurations to the set of benchmarks.
Table~\ref{bfc-experiments} shows the corresponding experimental results.

\begin{table}[h]
\begin{center}
  \begin{tabular}{ | r | p{7cm} | r | r | r | r | } %
    \hline
    \multicolumn{2}{|l|}{Configuration} & pos & neg & don't know & TO \\
    \hline
    C1 & State constraint over real numbers & 23 &  0 & 54 &  0 \\
    C2 & State constraint over integers        & 23 &  0 & 54 &  0 \\
    \hline
    C3 & C2 + TrapConditions                   & 23 &  0 & 54 &  0 \\
    C4 & C3 + SubnetTrapConditions         & 23 &  0 & 50 &  4 \\
    C5 & C4 + EmptyTrapConditions          & 23 &  0 & 46 &  8 \\
    \hline
    C6 & C1 + state space exploration \& TO 1\,min & 26 & 34 & 0 & 17 \\
    C7 & C1 + state space exploration \& TO 10\,min & 26 & 39 & 0 & 12 \\
    C8 & C1 + state space exploration \& TO 8\,h   & 26 & 45 & 0 &  6 \\
    \hline
    C10 & Method Invariant                  & 23 &  0 & 54 &  0 \\
    \hline
  \end{tabular}
\end{center}
\caption{Experimental results for the benchmarks of \bfc}
\label{bfc-experiments}
\end{table}

Observations:
\begin{itemize}
\item State constraints over integers or reals correctly determine that the 23 benchmarks in Appendix~\ref{bfc-experiments-c1} are positive.
\item Composing state space constraints with concrete state space
  exploration gives the following 3 additional positives.
\begin{verbatim}
cprover-PN/lamport
cprover-PN/newdekker
cprover-PN/peterson
\end{verbatim}
\end{itemize}

\newpage
\section{Benchmarks of \mist}

We experimented on the benchmarks available at \\
\ttt{https://github.com/pierreganty/mist/tree/master/examples/boundedPN}
and \\
\ttt{https://github.com/pierreganty/mist/tree/master/examples/PN}.
The set of benchmarks consists 27 benchmarks, 23 positive and 4 negative.
We applied 6 tool configurations to the set of benchmarks.
Table~\ref{mist-experiments} shows the corresponding experimental results.

\begin{table}[h]
\begin{center}
  \begin{tabular}{ | r | p{6cm} | r | r | r | r | }
    \hline
    \multicolumn{2}{|l|}{Configuration} & pos & neg & don't know & TO \\
    \hline
    C1 & State constraint with real numbers & 14 &  0 & 13 &  0 \\
    C2 & State constraint with integers     & 14 &  0 & 13 &  0 \\
    \hline
    C3 & C2 + TrapConditions               & 20 &  0 &  7 &  0 \\
    C4 & C3 + SubnetTrapConditions         & 21 &  0 &  6 &  0 \\
    C5 & C4 + EmptyTrapConditions          & 21 &  0 &  6 &  0 \\
    \hline
    C9 & C4 + state space exploration \& TO 4\,h & 21 &  2 & 0 &  4 \\
    \hline
    C10 & Method Invariant                  & 14 &  0 & 13 &  0 \\
    \hline
  \end{tabular}
\end{center}
\caption{Experimental results for the benchmarks of \mist}
\label{mist-experiments}
\end{table}

Observations:
\begin{itemize}
\item State constraints over integers or reals correctly determine that the 14 benchmarks in Appendix~\ref{mist-experiments-c1} are positive.
\item Composing state constraints with TrapConditions gives the following 6 additional positives.
\begin{verbatim}
PN/pingpong
PN/basicME
PN/multiME
boundedPN/lamport
boundedPN/newdekker
boundedPN/peterson
\end{verbatim}
\item Composing state constraints with TrapConditions and SubnetTrapConditions gives the additional positive \ttt{PN/manufacturing}.
  \item The remaining positive benchmarks are the following.
\begin{verbatim}
PN/extendedread-write
PN/extendedread-write-smallconsts
\end{verbatim}
\end{itemize}

\iffalse

\paragraph{Intermediate results}
Constraints $\mathcal C'(N, P)$ with
$Y_1 = \begin{pmatrix}target1 \end{pmatrix}$,
$Y_2 =\begin{pmatrix}nc & m & cs1 & cs2\end{pmatrix}$,
$A =\begin{pmatrix}0 & 0 & 1 & 1 \end{pmatrix}$ and
$b =\begin{pmatrix}2 \end{pmatrix}$:

\begin{align*}
  target1 - nc - m + cs1 \le& 0 \\ 
  target1 - nc - m + cs2 \le& 0 \\ 
  - target1 + nc + m - cs1 \le& 0 \\ 
  - target1 + nc + m - cs2 \le& 0 \\ 
2 \cdot nc + m <& 2 \cdot target1 \\
  target1 \ge& 0 \\
  nc \ge& 0 \\
  m \ge& 0 \\
  cs1 \ge& 0 \\
  cs2 \ge& 0
\end{align*}

Satisfying assignment is $Y_1 = \begin{pmatrix} 1 \end{pmatrix}$ and
$Y_2 = \begin{pmatrix}0 & 1 & 0 & 0 \end{pmatrix}$

Invariant $\text{Inv}(N, A) = (cs1 + cs2 + m \le 1)$

\paragraph{Technique}
\begin{align*}
  \mathcal C'(N, P) =& ( ( ( Y_1 \cdot A + Y_2) \cdot C \le 0) \land \\
& ((Y_1 \cdot A + Y_2) \cdot M_0 < Y_1 \cdot b) \land \\
& (Y_1 \ge 0) \land (Y_2 \ge 0)) \\
  \text{Inv}(N, A) =& ( ( Y_1 \cdot A + Y_2) \cdot M \le ( Y_1 \cdot A + Y_2) \cdot M_0)
\end{align*}
where $M_0$ is the initial marking, $C$ is the incidence matrix of the net and
$A\cdot M \ge b$ is equivalent to $\neg P$.

\paragraph{Method Invariant}
\mbox{ } \\
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C':=\mathcal C'(N, P)$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C')$};
  \node[print, right=of satc] (noinv) {No Invariant};
  \node[action, below=of satc] (inv) {$A_\theta:=\text{Model}(C')$\\
                                  	$I := \text{Inv}(N, A)$};
  \node[print, below=of inv] (printinv) {Invariant: $I$};
  \node[state, right=of noinv] (end1) {END};
  \node[state, below=of printinv] (end2) {END};

  \draw (begin) edge (c);
  \draw (c) edge (satc);
  \draw (satc) edge node[above]{NO} (noinv);
  \draw (satc) edge node[right]{YES} (inv);
  \draw (noinv) edge (end1);
  \draw (inv) edge (printinv);
  \draw (printinv) edge (end2);
\end{tikzpicture}

\paragraph{Example: Mutual exclusion algorithm in Delzanno2001}

Property: The number of threads in critical sections should be at most 1

* Petri net

Place m1 is marked in the net exactly if m1 is false in the algorithm.
Place m2 is marked in the net exactly if m2 is false in the algorithm.

\paragraph{Example: Petri net for mutual exclusion algorithm in Delzanno2001}
\mbox{ } \\
\begin{tikzpicture}[node distance=2cm]
  \node[blue place, tokens=2, label=above:nc] (nc) {};
  \node[place, tokens=1, label=right:m2, below of=nc, xshift=+0.75cm] (s2) {};
  \node[place, tokens=1, label=left:m1, below of=nc, xshift=-0.75cm] (s1) {};
  \node[red place, label=above:cs1, left of=s1] (cs1) {};
  \node[red place, label=above:cs2, right of=s2] (cs2) {};
  \node[transition, label=above:t1] (t1) at ($(s1)!0.5!(cs1) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s1)
	edge [pre] (s2)
	edge [post] (cs1)
	edge [post] (s2);
  \node[transition, label=above:t2] (t2) at ($(s2)!0.5!(cs2) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s2)
	edge [pre] (s1)
	edge [post] (cs2)
	edge [post] (s1);
  \node[transition, label=below:t3] (t3) at ($(s1)!0.5!(cs1) - (0,1cm)$) {}
	edge [pre] (cs1)
	edge [post] (s1)
	edge [post,bend left,min distance=3cm, out=115, in=115] (nc);
  \node[transition, label=below:t4] (t4) at ($(s2)!0.5!(cs2) - (0,1cm)$) {}
	edge [pre] (cs2)
	edge [post] (s2)
	edge [post,bend right,min distance=3cm, out=-115, in=-115] (nc);
\end{tikzpicture}
\begin{align*}
  P &= ( cs1 + cs2 \le 1 ) \\
  \neg P &= ( cs1 + cs2 \ge 2 )
\end{align*}

\paragraph{Intermediate results}
Constraints $\mathcal C(N)$ with
$M =\begin{pmatrix}nc & m1 & m2 & cs1 & cs2 \end{pmatrix}^T$,
$M_0 =\begin{pmatrix}2 & 1 & 1 & 0 & 0 \end{pmatrix}^T$,
$X =\begin{pmatrix}t1 & t2 & t3 & t4 \end{pmatrix}^T$ and
$C =\begin{pmatrix}
  -1 & -1 & 1 & 1 \\
  -1 &  0 & 1 & 0 \\
   0 & -1 & 0 & 1 \\
  1 & 0 & -1 & 0 \\
  0 & 1 & 0 & -1
\end{pmatrix}$:
\begin{align*}
  nc   =& 2 - t1 - t2 + t3 + t4 \\
  m1  =& 1 - t1 + t3 \\
  m2  =& 1 - t2 + t4 \\
  cs1 =& 0 + t1 - t3 \\
  cs2 =& 0 + t2 - t4 \\
  nc \ge& 0 \\
  m1 \ge& 0 \\
  m2 \ge& 0 \\
  cs1 \ge& 0 \\
  cs2 \ge& 0
\end{align*}

Satisfying assignment is
$M =\begin{pmatrix}0 & 0 & 0 & 1 & 1 \end{pmatrix}^T$ and
$X =\begin{pmatrix}1 & 1 & 0 & 0 \end{pmatrix}^T$.

A set satisfying the trap conditions is $S = \{m1, m2\}$.

The trap constraint is $\delta = (m1 + m2 \ge 1)$.

With the additional constraint, the set of constraints is unsatisfiable.

\paragraph{Technique trap conditions} For a petri net $N$ and an assignment $A$,
find a set $S$ that satisfies
\begin{enumerate}
  \item $S$ is a trap in the net $N$.
  \item $S$ is marked in the initial marking $M_0$.
  \item $S$ is unmarked in the assignment $A$.
\end{enumerate}
For such a set $S$, generate a constraint
$\delta = \left( \sum_{s \in S} s \ge 1 \right)$, ensuring the
trap is marked in any assignment.

\paragraph{Method Safety by Refinement}
\mbox{ } \\
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
  \node[action, below=of satc] (modelc) {$A:=\text{Model}(C \cup \{\neg P\})$\\
                               	$C_{\theta}:=\text{TrapConditions}(N, A)$};
  \node[decision, below=of modelc] (satctheta) {$\text{SAT}(C_\theta)$};
  \node[action, below=of satctheta] (modelctheta) {$A_\theta:=\text{Model}(C_\theta)$\\
                                    	$\delta:=\Delta(A_\theta)$\\
                                    	$C:=C \cup \{\delta\}$};
  \node[print, right=of satc] (yes) {YES};
  \node[print, right=of satctheta] (dontknow) {Don't know};
  \node[state, right=of yes] (end1) {END};
  \node[state, right=of dontknow] (end2) {END};

  \draw (begin) edge (c);
  \draw (c) edge coordinate[pos=.5] (edgein) (satc);
  \draw (satc) edge node[above]{NO} (yes);
  \draw (yes) edge (end1);
  \draw (satc) edge node[right]{YES} (modelc);
  \draw (modelc) edge (satctheta);
  \draw (satctheta) edge node[above]{NO} (dontknow);
  \draw (dontknow) edge (end2);
  \draw (satctheta) edge node[right]{YES} (modelctheta);
  \draw (modelctheta.south) -- ([yshift=-0.5cm] modelctheta.south)
  -| ([xshift=-2cm] modelctheta.west) |- (edgein);
\end{tikzpicture}

\section{Method Invariant by Refinement}

\paragraph{Example}
See example for method Safety by Refinement.

\paragraph{Intermediate results}
Constraints $\mathcal C'(N)$ with
$Y_1 = \begin{pmatrix}target1 \end{pmatrix}$,
$Y_2 = \begin{pmatrix}trap2 \end{pmatrix}$,
$Y_3 =\begin{pmatrix}nc & m & cs1 & cs2\end{pmatrix}$,
the trap from the trap refinement as
$D =\begin{pmatrix}0 & 1 & 1 & 0 & 0 \end{pmatrix}$,
$A =\begin{pmatrix}0 & 0 & 0 & 1 & 1 \end{pmatrix}$ and
$b =\begin{pmatrix}1 \end{pmatrix}$:

\begin{align*}
  target1 - trap1 - nc - m1 + cs1 \le& 0 \\ 
  target1 - trap1 - nc - m2 + cs2 \le& 0 \\ 
  - target1 + trap1 + nc + m1 - cs1 \le& 0 \\ 
  - target1 + trap1 + nc + m2 - cs2 \le& 0 \\ 
2 \cdot trap1 + 2 \cdot nc + m1 + m2 <& 2 \cdot target1 + trap1 \\
  target1 \ge& 0 \\
  trap1 \ge& 0 \\
  nc \ge& 0 \\
  m1 \ge& 0 \\
  m2 \ge& 0 \\
  cs1 \ge& 0 \\
  cs2 \ge& 0
\end{align*}

Satisfying assignment is
$Y_1 = \begin{pmatrix} 1 \end{pmatrix}$,
$Y_2 = \begin{pmatrix} 1 \end{pmatrix}$ and
$Y_3 = \begin{pmatrix}0 & 0 & 0 & 0 & 0 \end{pmatrix}$.

Invariant $\text{Inv}(N, D, P, A) =
((cs1 + cs2 + m1 + m2 \le 2) \land (m1 + m2 \ge 1))$.

\paragraph{Technique}
\begin{align*}
  \mathcal C'(N, D, P) =& ( ( ( Y_1 \cdot A + Y_2 \cdot D + Y_3) \cdot C \le 0) \land \\
& ((Y_1 \cdot A + Y_2 \cdot D + Y_3) \cdot M_0 < Y_1 \cdot b + Y_2 \cdot \mathbf{1}) \land \\
& (Y_1 \ge 0) \land (Y_2 \ge 0) \land (Y_3 \ge 0)) \\
  \text{Inv}(N, D, P, A) =& ( ( Y_1 \cdot A + Y_2 \cdot D + Y_3) \cdot M \le ( Y_1 \cdot A + Y_2 \cdot D + Y_3) \cdot M_0) \land ( D \cdot M \ge \mathbf{1} )
\end{align*}
where $M_0$ is the initial marking, $C$ is the incidence matrix of the net, $A\cdot M \ge b$ is equivalent to $\neg P$ and $D \cdot M \ge \mathbf{1}$ is equivalent to the trap constraints.

\paragraph{Method Invariant by Refinement}
\mbox{ } \\
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C:=\mathcal C(N)$\\
                                 	$D:=\{\}$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
  \node[action, below=of satc] (modelc) {$A:=\text{Model}(C \cup \{\neg P\})$\\
                                   	$C_{\theta}:=\text{TrapConditions}(N, A)$};
  \node[decision, below=of modelc] (satctheta) {$\text{SAT}(C_\theta)$};
  \node[action, below=of satctheta] (modelctheta) {$A_\theta:=\text{Model}(C_\theta)$\\
                                        	$\delta:=\Delta(A_\theta)$\\
                                        	$C:=C \cup \{\delta\}$\\
                                        	$D:=D \cup \{\delta\}$};
  \node[print, right=of satctheta] (noinv2) {No Invariant};
  \node[state, below=of noinv2] (end3) {END};

  \node[action, right=of satc] (cprime) {$C':=\mathcal C'(N, D, P)$};
  \node[decision, right=of cprime] (satcprime) {$\text{SAT}(C')$};
  \node[action, below=of satcprime] (inv) {$A_\theta:=\text{Model}(C')$\\
                                      	$I := \text{Inv}(N, D, P, A)$};
  \node[print, below=of inv] (printinv) {Invariant: $I$};
  \node[print, right=of satcprime] (noinv) {No Invariant};
  \node[state, below=of noinv] (end1) {END};
  \node[state, below=of printinv] (end2) {END};

  \draw (cprime) edge (satcprime);
  \draw (satcprime) edge node[above]{NO} (noinv);
  \draw (satcprime) edge node[right]{YES} (inv);
  \draw (inv) edge (printinv);
  \draw (noinv) edge (end1);
  \draw (printinv) edge (end2);

  \draw (begin) edge (c);
  \draw (c) edge coordinate[pos=.5] (edgein) (satc);
  \draw (satc) edge node[above]{NO} (cprime);
  \draw (satc) edge node[right]{YES} (modelc);
  \draw (modelc) edge (satctheta);
  \draw (satctheta) edge node[above]{NO} (noinv2);
  \draw (noinv2) edge (end3);
  \draw (satctheta) edge node[right]{YES} (modelctheta);
  \draw (modelctheta.south) -- ([yshift=-0.5cm] modelctheta.south)
  -| ([xshift=-2cm] modelctheta.west) |- (edgein);
\end{tikzpicture}

\fi

>>>>

\title{Report on a Safety Checker for Petri nets}
\author{Philipp Meyer \and Rusl\'{a}n Ledesma-Garza}
% \institute{Technische Universit\"at M\"unchen}
\date{Mon Nov 25 14:24:10 CET 2013}

@@

\begin{center}
  \begin{tikzpicture}
    \node[state] (begin) {BEGIN};
    \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
    \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
    \node[print, right=of satc] (yes) {N satisfies P};
    \node[print, below=of satc] (dontknow) {Don't know};
    \node[state, right=of yes] (end1) {END};
    \node[state, below=of dontknow] (end2) {END};

@@

  Subprocedure TrapConditions constructs trap conditions C_\theta corresponding to N and A.
  Subprocedure \Delta constructs refinement constraint \delta corresponding to A_\theta.
\end{verbatim}
\begin{center}
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
  \node[action, below=of satc] (modelc) {$A:=\text{Model}(C \cup \{\neg P\})$\\
                               	$C_{\theta}:=\text{TrapConditions}(N, A)$};
  \node[decision, below=of modelc] (satctheta) {$\text{SAT}(C_\theta)$};
  \node[action, below=of satctheta] (modelctheta) {$A_\theta:=\text{Model}(C_\theta)$\\
                                    	$\delta:=\Delta(A_\theta)$\\
                                    	$C:=C \cup \{\delta\}$};
  \node[print, right=of satc] (yes) {N satisfies P};
  \node[print, right=of satctheta] (dontknow) {Don't know};
  \node[state, right=of yes] (end1) {END};
  \node[state, right=of dontknow] (end2) {END};

@@

We experimented on the benchmarks \ttt{cprover-PN} and
\ttt{cprover\_software\_analysis} available at \\
\ttt{http://www.mpi-sws.org/\~{}jkloos/iic-experiments/cprover.zip}.
The set of benchmarks consists 77 benchmarks, 26 positive and 51 negative.
Table~\ref{bfc-experiments} shows the corresponding experimental results.

\begin{table}[h]
\begin{center}
  \begin{tabular}{ | p{7cm} | r | r | r | r | } %
    \hline
    Method & pos & neg & don't know & TO \\
    \hline
    Safety (over real numbers or naturals) & 23 &  0 & 54 &  0 \\
    Invariant                              & 23 &  0 & 54 &  0 \\
    Invariant Minimization                 & 23 &  0 & 54 &  0 \\
    Safety by Refinement                   & 23 &  0 & 54 &  0 \\
    \hline
  \end{tabular}
\end{center}
\caption{Experimental results for the benchmarks of \bfc}
\label{bfc-experiments}
\end{table}

\section{Benchmarks of \mist}

We experimented on the benchmarks available at \\
\ttt{https://github.com/pierreganty/mist/tree/master/examples/boundedPN}
and \\
\ttt{https://github.com/pierreganty/mist/tree/master/examples/PN}.
The set of benchmarks consists 27 benchmarks, 23 positive and 4 negative.
Table~\ref{mist-experiments} shows the corresponding experimental results.

\begin{table}[h]
\begin{center}
  \begin{tabular}{ | p{6cm} | r | r | r | r | }
    \hline
    Method & pos & neg & don't know & TO \\
    \hline
    Safety (over real numbers or naturals) & 14 &  0 & 13 &  0 \\
    Invariant                              & 14 &  0 & 13 &  0 \\
    Invariant Minimization                 & 14 &  0 & 13 &  0 \\
    Safety by Refinement                   & 20 &  0 &  7 &  0 \\
    \hline
  \end{tabular}
\end{center}
\caption{Experimental results for the benchmarks of \mist}
\label{mist-experiments}
\end{table}
